
```{r echo=TRUE, message=FALSE, warning=FALSE}
# install the necessary packages
install.packages(c("tidyverse", "dplyr", "readr", "tidyr"), repos = "https://cran.rstudio.com/")
library(tidyverse)
library(dplyr)
library(readr)
library(tidyr)

# import the dataset
water <- read.csv("F:\\R practice\\Dataset\\water_potability.csv")

# checking the structure and summary of the dataset
str(water)
summary(water)
```

#### Data Cleaning

```{r echo=TRUE, warning=FALSE}
# Checking for missing values if any in the dataset
colSums(is.na(water)) 
# It seems there are some missing values in columns 'ph', 'Sulfate', 'Trihalomethanes'.             

# Missing values were imputed with the mean.
water$Sulfate[is.na(water$Sulfate)] <- mean(water$Sulfate, na.rm = TRUE)

water$ph[is.na(water$ph)] <- mean(water$ph, na.rm = TRUE)

water$Trihalomethanes[is.na(water$Trihalomethanes)] <- mean(water$Trihalomethanes, na.rm = TRUE)

# Confirming the replacement of missing values with mean values.
any(is.na(water))

summary(water)
```

#### Analysis

```{r echo=TRUE}
# Replacing the values 0 & 1 in Potability column to "Potable" and "Not Potable".
water.filtered <- water %>% 
  mutate(Potability = ifelse(Potability == 1, "Potable", "Not Potable"))

# finding out the percentage of potable and not potable water.
perc.pot <- water.filtered %>% 
  group_by(Potability) %>% 
  summarise(Count = n()) %>% 
  mutate(Perc = (Count / sum(Count)) * 100)

 print(perc.pot) # Not Potable water is 61% and potable water is 39%
```

##### Digging deep into pH column

```{r echo=TRUE, warning=FALSE}
# finding the number of samples having PH scale of Not portable water
ph.filter <- water.filtered %>% 
  filter(ph < 6.5 | ph > 8.5)
  
ph.filter %>% 
  summarise(Count = n())

# Visualizing the count of pH scales within and outside the recommended limits.
water.filtered %>% 
  mutate(ph.category = ifelse(ph < 6.5 | ph > 8.5, "Outside Limit", "Within Limit")) %>% 
  group_by(ph.category) %>% 
  summarise(Count = n()) %>% 
  ggplot(aes(x = ph.category, y = Count, fill = ph.category))+
  geom_bar(stat = "identity", position = "dodge") +
  theme_minimal()+
  guides(fill = FALSE) +
  labs(title = "pH Scale Distribution: Within vs. Outside Recommended Limits")

# What is the average pH value across all water bodies?
mean(water.filtered$ph)

# What is the percentage of potable water bodies with a pH less than 7?
potable.water <- water.filtered %>% 
  filter(Potability == "Potable")

ph.7.less <- mean(potable.water$ph < 7) * 100

print(ph.7.less) # 42.09703 %
```

The ideal pH range for drinking water is between 6.5 and 8.5. This range is considered safe for human consumption. Despite a higher proportion of non-potable water in the sample, a majority of samples exhibit acceptable pH levels, with a count surpassing 1750. Based on this information, we can conclude that while there is a significant portion of non-potable water in the sample, a large number of samples still meet the pH standards for drinking water. This suggests that the pH level may not be the primary factor determining the potability of water in this sample.

```{r echo=TRUE}
# How does the distribution of hardness vary among potable and non-potable water?
water.filtered %>% 
  ggplot(aes(x = Potability, y = Hardness, fill = Potability))+
  geom_boxplot()+ theme_minimal()+
  guides(fill = FALSE)+
  labs(title = "Distribution of Hardness for Potable and Non-Potable Water")
```

The box plots for potable and non-potable water appear to have similar medians. The median hardness for both groups is around 200 mg/L. However, there is a difference in the interquartile ranges (IQRs) between the two groups. The IQR for portable water is slightly wider than the IQR for non-portable water, which suggests that the distribution of hardness for potable water is more spread out than the distribution of hardness for non-potable water. The wider IQR for potable water suggests that there are more extreme hardness values observed in potable water samples. This broader range of hardness levels could be due to various factors, such as differences in water treatment processes, natural variations in water sources, or contamination from external sources.

```{r echo=TRUE, warning=FALSE}
# What is the maximum and minimum total dissolved solids (TDS) recorded in potable water?

# method 1
summary(water.filtered$Solids)

# method 2
max.value <- max(water.filtered$Solids)
min.value <- min(water.filtered$Solids)

print(max.value)
print(min.value)

# method 3, combining both results to one output
max.min.TDS<- rbind(max.value, min.value)
rownames(max.min.TDS) <- c("Max TDS: ", "Min TDS: ")

print(max.min.TDS)

# Is there any relationship between chloramines and sulfate concentrations in the water?
correlation <- cor(water.filtered$Chloramines, water.filtered$Sulfate)

print(correlation)

ggplot(water.filtered, aes(x = Chloramines, y = Sulfate)) +
  geom_point(alpha = 0.5)+
  geom_smooth() +
  labs(title = "Scatterplot between Chloramines and Sulfates")

# digging deeper

ggplot(water.filtered, aes(x = Chloramines, y = Sulfate,)) +
  geom_point(alpha = 0.5, color = "brown") +
  geom_smooth() +
  facet_wrap(~ Potability) +
  theme_minimal() +
  labs(title = "Scatterplot between Chloramines and Sulfates")
```

Chloramines are primarily added to water as a disinfectant to kill harmful bacteria and pathogens. On the other hand, sulfates can enter water through natural sources, such as weathering of rocks and soil, or through anthropogenic activities like industrial waste discharge. The correlation coefficient of 0.02379109, which is very close to 0, indeed indicates an almost negligible linear relationship between chloramines and sulfate concentrations. This means that changes in the concentration of one variable are not accompanied by predictable changes in the concentration of the other variable. In simpler terms, there is no strong association between the levels of chloramines and sulfates in the water samples.

```{r echo=TRUE}
# Which variable(s) have the highest variance among all the features?
without.potability <- select(water.filtered, -Potability)

variances <- sapply(without.potability, var) 

print(variances)
# Solids are having the highest variance, with a variance of 7.688783e+07
```

Variances provide insights into the distribution and spread of data within each variable. They help assess the degree of variability, which is crucial for understanding the characteristics and nature of each feature in our dataset. Solids, Conductivity and Turbidity have more significant variability or dispersion in their data points compared to other variables. Higher variance suggests that data points are more spread out from the mean, indicating a wider range of values in the dataset.

```{r echo=TRUE}
# How many water bodies have sulfate levels above the recommended limit for drinking purposes?
sulfate.limit <- water.filtered %>% 
  filter(Sulfate > 250) # the permissible limit of sulfate level in drinking water is 250 mg/L

length(sulfate.limit$Sulfate) # 3218


# Is there a significant difference in the mean chloramine levels between potable and non-potable water?
potable.water <- water.filtered %>% 
  filter(Potability == "Potable")

non.potable <- water.filtered %>% 
  filter(Potability == "Not Potable")
  
# Perform t-test to compare mean Chloramine levels between potable and non-potable water
t.test.result <- t.test(potable.water$Chloramines, non.potable$Chloramines)

t.test.result

ggplot(water.filtered, aes(x = factor(Potability), y = Chloramines, fill = Potability)) +
  geom_boxplot() +
  theme_minimal()+
  labs(title = "Chloramine Levels between Potable and Non-potable Water", x = "Potability") +
  guides(fill = FALSE)
```

The box plot shows that the median chloramine level in potable water is higher than the median chloramine level in non-potable water. However, the interquartile range (IQR) for potable water is also wider, indicating that the distribution of chloramine levels in potable water is more spread out.

Based on the results of this Welch Two Sample t-test, with a p-value of 0.1856 greater than the typical significance level of 0.05, there isn't enough evidence to conclude a significant difference between the means of Chloramines in potable and non-potable water.

```{r echo=TRUE}
# What is the relationship between hardness and sulfate levels in the water?
cor.value <- cor(water.filtered$Hardness, water.filtered$Sulfate) #  -0.09276615

print(cor.value)

ggplot(water.filtered, aes(x = Hardness, y = Sulfate, color = Potability)) +
  geom_point(alpha = 0.5) +
  theme_minimal() +
  labs(title = "Scatterplot between Hardness and Sulfates in Water") 
  guides(fill = FALSE) 
```

The obtained correlation value of approximately -0.0928 indicates a weak negative correlation between hardness and sulfates in water. This means that as the hardness of water increases, the concentration of sulfates tends to decrease slightly. However, it is important to note that the correlation coefficient is very close to 0, which suggests that the relationship between the two variables is very weak.

<br>

#### Data Analysis and Insights:

##### ➣ pH Levels:

 - Majority of samples fell within the WHO recommended pH range of 6.5 to 8.5 for safe drinking         water.
 
 - Approximately 42% of potable water bodies had a pH less than 7, still within acceptable limits.

##### ➣ Hardness and Potability:

- Similar median hardness observed for potable and non-potable water, but wider range of values in     potable water suggests more extreme hardness levels.

##### ➣ Total Dissolved Solids (TDS):

- TDS levels varied considerably, with the highest variance among the features.

- Several water bodies recorded TDS values exceeding the recommended limit for drinking purposes.

##### ➣ Chloramines and Sulfate:

- Negligible linear relationship found between chloramines and sulfate concentrations.

- Potable water showed a slightly higher median chloramine level but with a wider spread.

##### ➣ Correlation between Hardness and Sulfate:

- A weak negative correlation exists between hardness and sulfate levels.

##### Key Takeaways:

- pH Compliance: A majority of samples met pH standards despite some deviation, implying satisfactory   acid-base balance.

- TDS and Potability: Elevated TDS levels observed in a considerable number of water bodies could      raise concerns regarding drinkability.

- Chloramines Comparison: No significant difference in mean chloramine levels between potable and      non-potable water, though potable water showed a wider spread.

- Correlation: Weak correlations imply minimal direct relationships between some parameters,           highlighting independent influences on water quality.

#### Conclusion:

The assessment provides insights into various water quality parameters, indicating areas of compliance and concern.

Further investigation is recommended for TDS levels surpassing drinking water limits, while understanding the nuances of factors influencing water quality is crucial for effective policymaking and ensuring safe drinking water for communities.
